PhD Thesis

### 2016-12-13 chap-intro.tex
Move PLDI14 contributions to intro

Therefore, in this thesis we propose techniques for optimizing collection
queries and executing them incrementally.\pg{Refine}

@@

In particular, this thesis is divided into two parts:
\begin{itemize}
\item we first describe, in \cref{sec:ch-aosd13}, work on optimizing collection queries by static program transformation.
\item in the remaining chapters, we describe work on incrementalizing programs
  by static program transformation. This thesis presents the first approach that
  handles higher-order programs by using program transformation; hence, while
  our main examples use collection queries, we phrase the work in terms of
  $\lambda$-calculi with unspecified primitives.
\end{itemize}
% Modern libraries offer high-level APIs for manipulating (in-memory) collections;
% equivalently, such an API is an embedded domain-specific language (EDSL).

>>>>

\section{This thesis}
To reduce the need for manual optimizations, in this thesis we propose
techniques for optimizing collection queries and executing them
incrementally.\pg{Refine}

@@

In particular, this thesis is divided into two parts:
\begin{itemize}
\item we first describe, in \cref{sec:ch-aosd13}, work on optimizing collection queries by static program transformation.
\item in the remaining chapters, we describe work on incrementalizing programs
  by static program transformation. This thesis presents the first approach that
  handles higher-order programs by using program transformation; hence, while
  our main examples use collection queries, we phrase the work in terms of
  $\lambda$-calculi with unspecified primitives.
\end{itemize}

The second part makes the following contributions:
\begin{itemize}
\item We present a novel mathematical theory of changes and derivatives, which is more
  general than other work in the field because changes are
  first-class entities, they are distinct from base values and
  they are defined also for functions (\cref{sec:1st-order-changes}).
  %KO: I think the next sentence cannot be understood at this point.
  %We introduce changes for complex types, defined compositionally.
%
\item We present the first approach to incremental computation for pure
  $\lambda$-calculi by a source-to-source transformation, $\DERIVE$, that
  requires no run-time support. The transformation produces an incremental
  program in the same language; all optimization techniques for the original
  program are applicable to the incremental program as well.
%KO: commented this out. I think the purity is not important enough
%to deserve another sentence here, since we only vaguely hint
%at "further research".
%Since our incremental programs use no impure features, they are
%especially amenable to further optimizations, making this approach
%very suitable for further research.
%
% KO: Let's have one bullet point per section. Also, a conjecture
% sounds like a rather weak contribution
%\item We argue that incrementalization is efficient on
%  \emph{self-maintainable programs}, and discuss how further research on
%  static or dynamic memoization can speed up a larger class of programs (\cref{sec:performance-cons}).
%  \pg{This contribution references text which is now commented
%    out. I believe the text should be brought back in.}
%
We prove that our incrementalizing transformation $\DERIVE$
is correct~(\cref{eq:correctness})
by a machine-checked formalization in Agda~\citep{agda-head}.
The proof gives insight into the definition of $\DERIVE$: we
first construct the derivative $\EvalInc{-}$ of the denotational
semantics of a simply-typed $\lambda$-calculus term, that is, its
\emph{change semantics}.
%
Then, we show that $\DERIVE$ is produced by erasing
$\EvalInc{-}$ to a simply-typed program (\cref{sec:correctness}).

\item While we focus mainly on the theory of changes
and derivatives, we also perform a performance case study.
We implement the derivation transformation in Scala,
with a plug-in architecture that can be extended with new base
types and primitives. We define a plugin with support for
different collection types and use the plugin to
incrementalize a variant of the MapReduce programming model~\citep{Lammel07}.
  Benchmarks show that on this program,
  incrementalization can reduce asymptotic complexity and can turn $O(n)$
  performance into $O(1)$, improving running time by over 4
  orders of magnitude on realistic inputs (\cref{sec:applying}).

\end{itemize}
% Modern libraries offer high-level APIs for manipulating (in-memory) collections;
% equivalently, such an API is an embedded domain-specific language (EDSL).
